Parallel Functional Arrays

### 2016-09-25 Paper/rebuttal.txt

Thanks to all the reviewers for detailed and useful feedback.

Review A:

- Thanks for your detailed comments, we will fix the issues you identified.

Review B:

- Thanks for the feedback on our presentation. We will make the workings of the data structure more clear.

- We will work on the clarity of our correctness sketch. Our argument is rely-guarantee, but we cite Birkedal et al for its extension to a higher order language using logical relations, which we depend on.

- Haskell's Diffarrays are an implementation of Bloss' idea of "trailers" (from "Update Analysis and the Efficient Implementation of Aggregates”). We cite the improved version by Chuang, but will add the earlier cite by Bloss and point out that Diffarrays use it. Diffarrays differ from ours in that: (1) reading in the past takes O(m) time, where m is the number of updates made (i.e. it is not even a function of the size), (2) it does not support concurrency, (3) there is no cost semantics.

Review C:

- Thanks for your detailed comments, we will fix all identified issues and ambiguities.

- O(n) vs O(n^2) time question: For every n operations on a leaf sequence, we only make a copy of the values in the leaf sequence. We don’t copy interior sequences, or the change logs stored at the leaf sequence. The aim is to keep the size of the change logs small - the newly copied sequence will have empty change logs. We will make this logic more clear in our paper.

- Exploring functional hashtables is a great idea.

Review D:

- Your observations are all valid. The sequential version of our data structure is conceptually simple, and the sequential analysis is easy, and not of significant interest from an algorithmic point of view. The main goal of our paper is to give a concurrent implementation, and analyze its complexity formally in a cost semantics. We present a way to do this without having to consider all possible interleavings of threads, and the analysis can generalize to other structures. From the data structures point of view, we are also interested in a wait-free concurrent implementation. We will identify our main contributions more clearly.

- Thanks for your detailed feedback on our experiments. Indeed, it is easy to construct cases where our data structure performs poorly (e.g. repeatedly setting interior nodes). Our aim was to give preliminary results that suggest our data structure works well if most operations are on leaf nodes, even if the data structure is used concurrently (the efficiency of leaf updates is not obvious due to potential overheads of the implementation). As you point out, more systematic testing should be done in future work.

Review E:

- Thanks for the detailed review, the full code (in SML and Java) is available at https://github.com/AnanyaKumar/PersistentArrays

- Indeed amortization would be a problem if applied to the span (depth), but this does not pose a problem for us since the copy can be done in parallel so the span of the copy is O(log(n)) in the worst case, not amortized (section 6.6). The amortization is only with respect to the work, which is valid since work adds. Given the work and depth, we rely on prior work to compute the costs for a machine with p processors (theorem 3.1). This analysis works even if the work is amortized.

- We will clarify the ambiguity about size vs capacity in Pusharray.new

- We will clarify that our algorithm only improves on the READ accesses to old versions (compared to Chuang’s array)

### 2016-09-25 Paper/costproof.tex
Reorganize repository

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% numbers       To obtain numeric citation style instead of author/year.

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{listings}

\newcommand{\cL}{{\cal L}}

\newcommand{\func}[1]{{\sc #1}}
\newcommand{\tabulate}{\func{tabulate}}
\newcommand{\get}{\func{get}}
\newcommand{\set}{\func{set}}
\newcommand{\tget}{\text{\get}}
\newcommand{\tset}{\text{\set}}
\newcommand{\ttab}{\text{\tabulate}}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\copyrightdoi{nnnnnnn.nnnnnnn}

% Uncomment the publication rights you want to use.
%\publicationrights{transferred}
%\publicationrights{licensed}     % this is the default
%\publicationrights{author-pays}

% \titlebanner{banner above paper title}        % These are ignored unless
% \preprintfooter{short description of paper}   % 'preprint' option specified.

\title{A New Approach for Parallel Functional Arrays}

\authorinfo{Ananya Kumar}
           {Carnegie Mellon University}
           {ananyak@andrew.cmu.edu}
\authorinfo{Guy Blelloch}
           {Carnegie Mellon University}
           {blelloch@cs.cmu.edu}
\authorinfo{Robert Harper}
           {Carnegie Mellon University}
           {rwh@cs.cmu.edu}
\maketitle

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\begin{abstract}
In this paper we introduce a $O(1)$ wait-free, parallel, functional array, that allows $O(1)$ reads and writes to the most recent version of the array. We describe the cost dynamics and sketch out a provable implementation. We show favorable benchmarks comparing our functional arrays with regular arrays in Java.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\keywords
array, parallel, cost semantics

\input{intro}

\section{Dynamics}

We use a standard applicative-order source language defined as follows:
$$e = x \; | \; c  \; | \; \lambda x.e \; 
| \; e_1 e_2 \; 
| \; (e_1, e_2) \; 
| \; (e_1 || e_2) \; 
| \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$

The constants $c$ contains the usual arithmetic types, such as the
natural numbers and numerical functions.  The expression $(e_1||e_2)$
indicates that the two expressions can run in parallel, returning a
pair when both expressions are fully evaluated, and $(e_1,e_2)$
generates a pair sequentially.  The terminal (fully evaluated) values
are lambda expressions, as well as pairs and arrays of terminal
values, and are denoted by the judgement val.   Our value semantics are given by the judgement:
\[e \Downarrow v\]
indicating that the expression $e$ evaluates to $v$.  The rules for
function application and parallel pair are given as examples, and
other rules are standard.
$$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
$$\frac{f \Downarrow \lambda x . e \;\;\; y \Downarrow v \;\;\; [v/x]e \Downarrow v'}{f \; y \Downarrow v'} \text{ (function-app)}$$

The language also includes three functions for working with arrays: \tabulate{}, \get{}, and \set{}. The dynamics for \tabulate{} is given below:
$$\frac{f(1) \Downarrow v_1, ..., f(n) \Downarrow v_n}{\text{\tabulate{}} \; f \; n \Downarrow [1 \mapsto v_1, ..., n \mapsto v_n]}$$

\get$(A,i)$ returns the value at the $i$th index of $A$. \set$(A,i,v)$ returns a new array where the value at index $i$ is $v$ and the value at all other indices is the same as in $A$. The evaluation of a simple program in our language is shown below:\\

$(\lambda A. \; \tget(A, 2) + \tget(A,3)) \tset(\ttab(\lambda i . \; i, 5), 2, 10)$

$(\lambda A. \; \tget(A, 2) + \tget(A,3)) \tset([1,2,3,4,5], 2, 10)$

$(\lambda A. \; \tget(A, 2) + \tget(A,3)) [1,10,3,4,5]$

$\tget([1,10,3,4,5], 2) + \tget([1,10,3,4,5],3) $

$10+3$

$13$

\section{Cost Dynamics}

The dynamics given in the previous section are functional. However, the costs of get and set depend on whether the array is a leaf or internal array. To deal with this, we thread a store through the cost dynamics to keep track of whether each array is a leaf or internal array.

The cost dynamics of fork-join pose an additional challenge. An array can be used concurrently by multiple threads and the costs could depend on the order in which the instructions are interleaved. Figure \ref{fig:fork_join_intuition} shows 3 ways a leaf array $A$ might be used. In particular, one side of the fork might call get on $A$ while the other side calls set (see the center panel). Calls to get on the left branch would be $O(1)$ if and only if they execute before the call to set on the right branch.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.3]{fork_join_intuition}
\nocaptionrule \caption{Different ways an array might be used in a fork join}
\label{fig:fork_join_intuition}
\end{figure}

We cannot make any assumptions about how the instructions in different branches of a fork are interleaved, so we give the worst case time complexity over all possible interleavings. We first compute the work of each fork separately. Then for arrays that are used in both forks, we add additional work at the join point. To compute the additional work at the join point, we keep track of the number of gets that cost $O(1)$ on each side of the fork. If the array was modified on the other side of the fork, then the modification might have happened before the get, so we need to charge additional work for the get.

Our cost dynamics are defined by the following judgement, where $\delta$ is the store, $e$ is the expression, $\delta'$ is the new store, $v$ is the value returned by evaluating $e$, $w$ is the work, and $s$ is the span.
$$\delta; e \boldsymbol\Downarrow \delta'; v; w; s$$

We give the cost dynamics for arbitrary expressions in our language:

$$\frac{}{\delta; c \Downarrow \delta; c; 1; 1} \text{ (constants)}$$

$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta'; \lambda x . e; w_1 ; s_1 \;\;\; \delta'; e_2 \Downarrow \delta'';v; w_2 ; s_2 \;\;\; \\ \delta''; [v/x]e \Downarrow \delta'''; v'; w_3 ; s_3\end{gathered}}{\delta; e_1 \; e_2 \Downarrow \delta'''; v'; 1+w_1+w_2+w_3; 1+s_1+s_2+s_3} \text{ (func-app)}$$

$$\frac{\delta; e_1 \Downarrow \delta'; true; w_1; s_1 \;\;\; \delta'; e_2 \Downarrow \delta''; v; w_2; s_2}{\delta; \text{if } e_1 \text{ then } e_2 \text{ else } e_3 \Downarrow \delta''; v; 1+w_1+w_2; 1+s_1+s_2} \text{ (if-true)}$$

$$\frac{\delta; e_1 \Downarrow \delta'; false; w_1; s_1 \;\;\; \delta'; e_3 \Downarrow \delta''; v; w_2; s_2}{\delta; \text{if } e_1 \text{ then } e_2 \text{ else } e_3 \Downarrow \delta''; v; 1+w_1+w_2; 1+s_1+s_2} \text{ (if-false)}$$

Arrays are represented by $(l, V)$ where $l$ is a label that uniquely identifies the array, and $V$ contains the values in the array. $V[i]$ is the $i^{\text{th}}$ value in the array. If \set{} has been called on an array, the array is an interior array (represented by $-$), otherwise the array is a leaf array (represented by $+$).

Array functions have different costs depending on whether the array is a leaf or interior array. We thread a store $\delta$ through each operation. $\delta$ maps array labels $l$ to $(+/-, c)$, where $c$ represents the number of cheap \get{}s on the array. $L(\delta)$ denotes the set of labels in the store $\delta$. Let $f(n)$ be the work of calling \get{} on an interior array of size $n$ and $g(n)$ be the work of calling \set{} on an interior array of size $n$.

The cost dynamics for the functions on a functional array of size $n$ are:
$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; v \text{ val} \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' \not\in L(\delta)}{\delta; \tset \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$

$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; v \text{ val} \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' \not\in L(\delta)}{\delta; \tset \; A \; i \; v \Downarrow \delta[l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$

$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta_1; A; w_1; s_1 \;\;\; \delta_1; e_2 \Downarrow \delta_2; i; w_2; s_2 \;\;\; \\ \delta_2; e_3 \Downarrow \delta_3; v; w_3; s_3 \;\;\; \delta_3; \tset \; A \; i \; v \Downarrow \delta'; A'; w'; s' \end{gathered}}{\delta; \tset \; e_1 \; e_2 \; e_3 \Downarrow \delta'; A'; w_1+w_2+w_3+w'; s_1+s_2+s_3+s'} \text{ (set-eval)}$$

$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$

$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; \delta[l \mapsto (-, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta; V[i]; f(n); 1} \text{  (get-interior)}$$

$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta_1; A; w_1; s_1 \;\;\; \delta_1; e_2 \Downarrow \delta_2; i; w_2; s_2 \;\;\; \\ \delta_2; \tget \; A \; i \Downarrow \delta'; v'; w'; s' \end{gathered}}{\delta; \tget \; A \; i \Downarrow \delta'; v'; w_1 + w_2 + w'; s_1 + s_2 + s'}\text{ (get-eval)}$$

$$\frac{\begin{gathered}\delta, e_L \Downarrow \delta_L; v_L; w_L, s_L \;\;\; \delta, e_R \Downarrow \delta_R; v_R; w_R, s_R \;\;\; \\ (L(\delta_L) \setminus L(\delta)) \cap (L(\delta_R) \setminus L(\delta)) = \emptyset\end{gathered}}{\delta; (e_L || e_R) \Downarrow \delta'; (v_L || v_R); 1 + w_L + w_R + w'; \max(s_L, s_R)} \text{  (fork-join)}$$

\set{} creates a new label and array and modifies the store to indicate that the newly created array is a leaf array, and the array \set{} was called on has become an interior array. The work for \set{} is higher for interior arrays. \get{} returns the value at the desired index, if the array is a leaf then the operation costs 1 unit of work but we increment the counter of cheap \get{}s in the store.

The fork-join cost semantics are the most interesting. Suppose we have an expression $e_L || e_R$ and
\[ \delta, e_L \Downarrow \delta_L; v_L; w_L, s_L \]
\[ \delta, e_R \Downarrow \delta_R; v_R; w_R, s_R \]

Further, suppose that $(\delta_L \setminus \delta) \cap (\delta_R \setminus \delta) = \emptyset$ (the new labels produced on both sides of the fork-join do not conflict). Consider an array $A = (l, V)$ with $\delta[l] = (s,c)$, $\delta_L[l] = (s_L, c + c_L)$, $\delta_R[l] = (s_R, c + c_R)$. When multiplying two signs or multiplying a sign with an integer, consider + to be 1 and - to be 0.

\func{Combine} describes how to combine the values in the store on both sides of the fork-join. The array is a leaf array iff it is a leaf array on both sides of the fork-join. Cheap \get{}s on one side of the fork remain cheap iff there were no calls to \set{} on the other side of the fork.
\begin{equation*}
  \begin{aligned}
   \text{\func{combine}}&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
   &(s_L s_R, c + s_R c_L + s_L c_R)
  \end{aligned}
\end{equation*}
$$\delta' = \; (\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \bigcup_{l \in L(\delta)} [l \mapsto \text{\func{combine}}(\delta[l], \delta_L[l], \delta_R[l])]$$
\func{Extrawork} gives the additional cost incurred if an array of size $n$ was modified on either side of a fork-join. If the array was an interior array before the fork-join, then all functions incurred their maximal cost and there is no additional cost. Otherwise, cheap \get{}s on one side of the fork are charged $f(n)$ work iff the other side of the fork called \set{}. If both sides of the fork called \set{}, then one of the \set{}s came first and has work $1$, the subsequent \set{} has work $g(n)$.
\begin{equation*}
  \begin{aligned}
  \text{\func{extra}}&\text{\func{work}}(l, (s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
  & ((\neg s_R) c_L  + (\neg s_L) c_R)(f(size(l))-1) + \\
  &s(\neg s_R) (\neg s_L) (g(size(l))-1)
 \end{aligned}
\end{equation*}
$$w' = \sum_{l \in L(\delta), \delta[l \mapsto (+, c)]} \text{\func{extrawork}}(l, \delta[l], \delta_L[l], \delta_R[l])$$
Then, the fork-join cost semantics are:
$$\delta; (e_L || e_R) \Downarrow \delta'; (v_L || v_R); 1 + w_L + w_R + w'; \max(s_L, s_R)$$

\tabulate{} involves multiple function calls in parallel, so its cost semantics use \func{combine} and \func{extrawork} multiple times. We omit discussion of \tabulate{} here and in the rest of the paper because it involves messy details that are not relevant to the goals of this paper.

\section{Analysis}

\subsection{Interleaved Structural Dynamics}

To show that our implementation is correct and to validate our cost dynamics, we introduce interleaved structural dynamics for the source language. The structural dynamics describe the possible steps a program in our source language can take and are given by the following judgement, where $S$ is the store, $e$ is the expression, $S'$ is the resulting store, $e'$ is the resulting expression after a step has been taken, and $w$ is the work.

$$S, e \to S', e', w$$

For now we let the span of each step be 1 (we account for the spans later on).

As before, arrays are represented by $(l, V)$ where $l$ is a unique label and $V$ contains the values in the array. However, the store $S$ maps labels only to $+$ (leaf) or $-$ (interior). $L(S)$ denotes the set of labels in the store $S$. \func{size}$(l)$ denotes the size of the array referenced by label $l$. The judgement \func{val} describes terminal values. The structural dynamics are given as follows:
$$\frac{e_2 \; \text{val}}{S, (\lambda x.e)e_2 \to S, [e_2/x]e, 1} \text{ (app-sub)}$$
$$\frac{S, e_2 \to S', e_2', w}{S, (\lambda x.e) e_2 \to S', (\lambda x.e) e_2', w} \text{ (app-arg)}$$
$$\frac{S, e_1 \to S', e_1', w}{S, e_1 e_2 \to S', e_1' e_2, w} \text{ (app-func)}$$
$$\frac{}{S, \text{if } true \text{ then } e_2 \text{ else } e_3 \to S, e_2, 1} \text{ (if-true)}$$
$$\frac{}{S, \text{if } false \text{ then } e_2 \text{ else } e_3 \to S, e_3, 1}  \text{ (if-false)}$$
$$\frac{S, e_1 \to S', e_1', w}{S, \text{if } e_1 \text{ then } e_2 \text{ else } e_3 \to S', \text{if } e_1' \text{ then } e_2 \text{ else } e_3, w} \text{ (if-cond)}$$
$$\frac{A = (l, V) \;\;\; S[l] = + \;\;\; i \text{ val}}{S, get(A, i) \to S, V[i], 1} \text{ (get-leaf)}$$
$$\frac{A = (l, V) \;\;\; S[l] = -  \;\;\; i \text{ val}}{S, get(A, i) \to S, V[i], f(size(l))} \text{ (get-interior)}$$
$$\frac{A = (l, V) \;\;\; S, e_2 \to S', e_2', w}{S, get(A, e_2) \to S', get(A, e_2'), w} \text{ (get-index)}$$
$$\frac{S, e_1 \to S', e_1', w}{S, get(e_1, e_2) \to S', get(e_1', e_2), w} \text{ (get-array)}$$
$$\frac{A = (l, V) \;\;\; S[l] = + \;\;\;  l' \not\in L(S) \;\;\; i \text{ val} \;\;\; v \text{ val}}{S, set(A, i, v) \to S[l \mapsto -], (l', V[i \mapsto v]), 1} \text{ (set-leaf)}$$
$$\frac{A = (l, V) \;\;\; S[l] = - \;\;\;  l' \not\in L(S) \;\;\; i \text{ val} \;\;\; v \text{ val}}{S, set(A, i, v) \to S, (l', V[i \mapsto v]), g(size(l))} \text{ (set-interior)}$$
$$\frac{A = (l, V) \;\;\; i \text{ val} \;\;\; S, e_3 \to S', e_3', w}{S, set(A, i, e_3) \to S', set(A,i,e_3'), w} \text{ (set-val)}$$
$$\frac{A = (l, V) \;\;\; S, e_2 \to S', e_2', w}{S, set(A, e_2, e_3) \to S', set(A,e_2',e_3), w} \text{ (set-index)}$$
$$\frac{e_1 \to S', e_1', w}{S, set(e_1, e_2, e_3) \to S', set(e_1',e_2,e_3), w} \text{ (set-array)}$$

The structural dynamics for fork-join is non-deterministic since either side of the fork can take a step.
$$\frac{S, e_1 \to S', e_1', w}{S, e_1 || e_2 \to S', e_1' || e_2, w} \text{ (fork-left)}$$
$$\frac{S, e_2 \to S', e_2', w}{S, e_1 || e_2 \to S', e_1 || e_2', w} \text{ (fork-right)}$$

The structural dynamics for tabulate is also non-deterministic.

\begin{definition}
A \emph{transition sequence} $T$ is a sequence of states $[(S_0, e_0), ..., (S_n, e_n)]$ and costs $[w_1, ..., w_n]$ s.t. for all $0 \leq i < n$, $S_i, e_i \to S_{i+1}, e_{i+1}, w_{i+1}$ . We say that $T$ takes $S_0, e_0$ to $S_n, e_n$ and has length $n$ and denote this by $S_0, e_0 \to^n S_n, e_n$.
\end{definition}

\begin{definition}
We say that $T$ is \emph{maximal} if there does not exist $S', e', w'$ s.t. $S_n, e_n \to S', e', w'$.
\end{definition}

\begin{definition}
A \emph{transition subsequence} $T_{i, j}$ of $T$ is given by the sequence of states $[(S_i, e_i), ..., (S_j, e_j)]$ and costs $[w_{i+1}, ..., w_j]$. Note that $T = T_{0,n}$.
\end{definition}

\begin{definition}
The \emph{work of a transition sequence} $T$ is given by:
$$W(T) = \sum_{i=1}^n w_i$$
Different transition sequences starting and ending at the same states may have different work.
\end{definition}

\subsection{Sequential Cost Analysis}

\begin{definition}
Consider a transition sequence $T$. We say that $S_i, e_i \to S_{i+1}, e_{i+1}, w_{i+1}$ is a get on $l$ if the step involves evaluating the get function on array $(l, V)$. We say it is a \emph{cheap get} on $l$ if $w_{i+1} = 1$. The number of cheap gets on $l$ in $T$ is denoted by $SC_l(T)$.
\end{definition}

\begin{definition}
Unlike the store in the evaluational dynamics, the store in the structural dynamics only stores whether each array is $+$ (leaf) or $-$ (interior). $sign(\delta)$ takes a store that maps $l \mapsto (s,c)$ and returns a store which maps $l \mapsto s$.
\end{definition}

\begin{definition}
Suppose that $\delta, e \Downarrow \delta', e', w', s'$. We define the number of cheap gets in going from $\delta$ to $\delta'$ in the evaluational dynamics as follows. Suppose $l \mapsto (s', c') \in \delta'$. If $l \mapsto (s,c) \in \delta$ then $EC_l(\delta, \delta') = c' - c$ and if $l \not\in \delta$ then $EC_l(\delta, \delta') = c'$. If $l \not\in \delta'$ then $EC_l(\delta, \delta') = 0$.
\end{definition}

\begin{definition}
A \emph{relabeling} of $L_1$ is a bijective function $R$ from label sets $L_1 \to L_2$. $R$ can be used to relabel stores and expressions. $R(\delta) = \{ R(l) \mapsto e \; | \; l \mapsto e \in \delta \wedge l \in L_1  \} \cup \{ l \mapsto e \; | \; l \mapsto e \in \delta \wedge l \not\in L_1  \}$. In other words, $R$ relabels some of the labels in $\delta$. Similarly, $R(e)$ returns an expression $e'$ where each occurrence of a label $l$ in $e$ is substituted by $R(l)$.
\end{definition}

@@

\begin{proof}
By casing on each rule in the structural dynamics, and noting that transitions of expressions never depend on the values in the store.
\end{proof}

@@

and consider arbitrary maximal transition sequence T taking 
$$S_0 = sign(\delta), e_0 \to^n S_n, e_n$$
å
The following hold:
\begin{enumerate}
\item For some relabeling $R$ of $L(\delta_E) \setminus L(\delta)$, $S_n = sign(R(\delta_E))$ and $e_n = R(e_E)$.
\item For all $l \in \delta_E$, $EC_l(\delta, \delta_E) \leq SC_{R(l)}(T)$.
\item The sum of work and cheap gets is conserved: \begin{align*}
w_E + &\sum_{l \in L(\delta_E)} EC_l(\delta, \delta_E) (f(size(l))-1) =\\
 &W(T) + \sum_{l \in L(\delta_E)} SC_{R(l)}(T) (f(size(R(l))) - 1)
 \end{align*}
\end{enumerate}
\end{theorem}

@@

\noindent \textbf{Case get-leaf}: Suppose $e_0 = get(e_L, e_R)$ and
$$\delta, e_L \Downarrow \delta_L, e_L', w_L$$
$$\delta_L, e_R \Downarrow \delta_R, e_R', w_R$$
$$\delta_R, get(e_L', e_R') \Downarrow \delta_E, e_E, 1$$
which gives us
$$\delta, get(e_L, e_R) \Downarrow \delta_E, e_E, w_E$$
with $w_E = w_L + w_R + 1$.

Note that the structural dynamics first step the left argument to get, then the right argument, and finally evaluate the get. So there exists $m$ such that $T_{0,m}$ involves stepping the left argument, $T_{m,n-1}$ involves stepping the right argument, and $T_{n-1,n}$ involves evaluating the get.

@@

\textbf{Part 2}: From lemma \ref{relabeling_lemma}, the length of the shortest derivation for the relabeled evaluational dynamics does not change, so we can apply the inductive hypothesis. Applying the IH to $T_{0,m}$, for all $l \in \delta_L$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m}') = SC_l(T_{0,m})$. If $l \in \delta_R$ but $l \not\in \delta_L$ then $EC_l(\delta, \delta_L) = SC_l(T_{0,m}) = 0$ (intuitively the label did not exist and so did not have any cheap gets), so in particular, for all $l \in \delta_R$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m})$. Applying the IH to $T_{m,n-1}$ we get that for all $l \in \delta_R$, $EC_l(\delta_L, \delta_R) \leq SC(T_{m,n-1})$.

@@

\textbf{Part 3}: The third part of the theorem follows by similarly applying the IH on $T_{0,m}$ and $T_{m,n-1}$, and noting that $W(T) = W(T_{0,m}) + W(T_{m,n-1}) + 1$.
\\

\noindent \textbf{Case fork-join}: Suppose $e_0 = (e_L || e_R)$ and
$$\delta, e_L \Downarrow \delta_L, v_L, w_L$$
$$\delta, e_R \Downarrow \delta_R, v_R, w_R$$
which gives us
$$\delta, (e_L || e_R) \Downarrow \delta', (v_1 || v_2), 1 + w_L + w_R + w'$$
with $w'$ and $\delta'$ defined in terms of \func{extrawork} and \func{combine} as given in the evaluational dynamics.

@@

By inducting on the product $mk$, we can show that the new labels produced in $T_L$ and $T_R$ do not overlap, that is: $(L(S^L_m) \setminus L(S_0)) \cap (L(S^R_k) \setminus L(S_0)) = \emptyset$. (Note to self: Should I sketch out this induction?). Compose the relabelings to get relabeling R.

Applying lemma $\ref{relabeling_lemma}$ and the fork join rule, we get,
$$\delta, R(e_L) \Downarrow R(\delta_L), R(v_L), w_L$$
$$\delta, R(e_R) \Downarrow R(\delta_R), R(v_R), w_R$$
which gives us
$$\delta, R((e_L || e_R)) \Downarrow R(\delta'), R((v_1 || v_2)), 1 + w_L + w_R + w'$$

>>>>

\section{Cost Proofs}
\label{sec:costproof}

We show that the work computed by the evaluational dynamics is a tight upper bound for the work in the interleaved structural dynamics. Proofs for the span bounds are omitted because they are standard (the parallel structural dynamics is deterministic).

\begin{definition}
Consider a transition sequence $T$. We say that $S_i, e_i \to S_{i+1}, e_{i+1}, w_{i+1}$ is a \get{} on $l$ if the step involves evaluating the \get{} function on structure $(l, V)$. We say it is a \emph{cheap \get{}} on $l$ if $w_{i+1} = g_l(l)$. The number of cheap \get{}s on $l$ in $T$ is denoted by $SC_l(T)$.
\end{definition}

\begin{definition}
Unlike the store in the evaluational dynamics, the store in the structural dynamics only stores whether each structure is $+$ (leaf) or $-$ (interior). $sign(\delta)$ takes a store that maps $l \mapsto (s,c)$ and returns a store which maps $l \mapsto s$.
\end{definition}

\begin{definition}
Suppose that $\delta, e \Downarrow \delta', e', w', s'$. We define the number of cheap \get{}s in going from $\delta$ to $\delta'$ in the evaluational dynamics as follows. Suppose $l \mapsto (s', c') \in \delta'$. If $l \mapsto (s,c) \in \delta$ then $EC_l(\delta, \delta') = c' - c$ and if $l \not\in L(\delta)$ then $EC_l(\delta, \delta') = c'$. If $l \not\in L(\delta')$ then $EC_l(\delta, \delta') = 0$.
\end{definition}

\begin{definition}
A \emph{relabeling} of $L_1$ is a bijective function $R$ from label sets $L_1 \to L_2$. $R$ can be used to relabel stores and expressions. $R(\delta) = \{ R(l) \mapsto e \; | \; l \mapsto e \in \delta \wedge l \in L_1  \} \cup \{ l \mapsto e \; | \; l \mapsto e \in \delta \wedge l \not\in L_1  \}$. In other words, $R$ relabels some of the labels in $\delta$. Similarly, $R(e)$ returns an expression $e'$ where each occurrence of a label $l \in L_1$ in $e$ is substituted by $R(l)$.
\end{definition}

@@

\begin{proof}
By casing on each rule in the structural dynamics, and noting that transitions of expressions never depend on the values in the store (the dynamics are purely functional).
\end{proof}

@@

and consider arbitrary maximal transition sequence T taking 
$$S_0 = sign(\delta), e_0 \to^n S_n, e_n$$
å
The following hold:
\begin{enumerate}
\item For some relabeling $R$ of $L(\delta_E) \setminus L(\delta)$, $S_n = sign(R(\delta_E))$ and $e_n = R(e_E)$.
\item For all $l \in \delta_E$, $EC_l(\delta, \delta_E) \leq SC_{R(l)}(T)$.
\item The sum of work and cheap \get{}s is conserved: \begin{align*}
w_E + &\sum_{l \in L(\delta_E)} EC_l(\delta, \delta_E) (g_i(l)-g_l(l)) =\\
 &W(T) + \sum_{l \in L(\delta_E)} SC_{R(l)}(T) (g_i(R(l)) - g_l(R(l)))
 \end{align*}
\end{enumerate}
\end{theorem}

@@

\noindent \textbf{Case get-leaf}: Suppose $e_0 = \tget{}(e_L, e_R)$ and
$$\delta, e_L \Downarrow \delta_L, e_L', w_L$$
$$\delta_L, e_R \Downarrow \delta_R, e_R', w_R$$
$$\delta_R, \tget{}(e_L', e_R') \Downarrow \delta_E, e_E, g_l(l')$$
where $e_L' = (l', V)$ for some $V$, which gives us
$$\delta, \tget{}(e_L, e_R) \Downarrow \delta_E, e_E, w_E$$
with $w_E = w_L + w_R + g_l(l')$.

Note that the structural dynamics first step the left argument to \get{}, then the right argument, and finally evaluate the \get{}. So there exists $m$ such that $T_{0,m}$ involves stepping the left argument, $T_{m,n-1}$ involves stepping the right argument, and $T_{n-1,n}$ involves evaluating the \get{}.

@@

\textbf{Part 2}: From lemma \ref{relabeling_lemma}, the length of the shortest derivation for the relabeled evaluational dynamics does not change, so we can apply the inductive hypothesis even after the relabeling. Applying the IH to $T_{0,m}$, for all $l \in \delta_L$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m}') = SC_l(T_{0,m})$. If $l \in \delta_R$ but $l \not\in \delta_L$ then $EC_l(\delta, \delta_L) = SC_l(T_{0,m}) = 0$ (intuitively the label did not exist and so did not have any cheap gets), so in particular, for all $l \in \delta_R$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m})$. Applying the IH to $T_{m,n-1}$ we get that for all $l \in \delta_R$, $EC_l(\delta_L, \delta_R) \leq SC(T_{m,n-1})$.

@@

\textbf{Part 3}: The third part of the theorem follows by similarly applying the IH on $T_{0,m}$ and $T_{m,n-1}$, and noting that $W(T) = W(T_{0,m}) + W(T_{m,n-1}) + g_l(l')$.
\\

\noindent \textbf{Case fork-join}: Suppose $e_0 = (e_L || e_R)$ and
$$\delta, e_L \Downarrow \delta_L, v_L, w_L$$
$$\delta, e_R \Downarrow \delta_R, v_R, w_R$$
which gives us
$$\delta, (e_L || e_R) \Downarrow \delta', (v_1 || v_2), w_L + w_R + w'$$
with $w'$ and $\delta'$ defined in terms of \func{extrawork} and \func{combine} as given in the evaluational dynamics.

@@

By inducting on the product $mk$, we can show that the new labels produced in $T_L$ and $T_R$ do not overlap, that is: $(L(S^L_m) \setminus L(S_0)) \cap (L(S^R_k) \setminus L(S_0)) = \emptyset$. Compose the relabelings to get relabeling R.

Applying lemma $\ref{relabeling_lemma}$ and the fork join rule, we get,
$$\delta, R(e_L) \Downarrow R(\delta_L), R(v_L), w_L$$
$$\delta, R(e_R) \Downarrow R(\delta_R), R(v_R), w_R$$
which gives us
$$\delta, R((e_L || e_R)) \Downarrow R(\delta'), R((v_1 || v_2)), w_L + w_R + w'$$
